---
title: "Modeling Depth With A General Additive Model"
output: html_notebook
---
The previous OLS and Lasso models only explained a small proportion of the observed event depths, as evidenced by the linear relationship between depths and residuals. We anticipate that a model with more flexibility will improve our predictive ability. For this task, I implement a General Additive Model (GAM) using natural splines. 
These natural splines are functions that allow more flexibility than requiring the relationship between parameters and output to be linear. Additionally, the additive property of the GAM means that the model maintains a fair level of interpretability, as we can examine each function comprising it and see how depth responds to changes in each feature (while holding the rest constant). 


```{r}
#loading libraries and reading in data

library(readr)
library(Metrics)
library(glmnet)
library(boot)
library(gam)
library(splines)
library(mltools)
library(data.table)
library(dplyr)
library(ggplot2)


input.train.file = "C:/Users/riley/Seismo_Direc/cleaned_training_df.csv"
input.test.file = "C:/Users/riley/Seismo_Direc/cleaned_testing_df.csv"

train <- read_csv(input.train.file)
test <- read_csv(input.test.file)
```

Model Fitting

When using natural splines, we don't know beforehand the number of degrees of freedom (our hyper-parameters) for each parameter. To estimate these hyper-parameters, we perform a four-dimensional grid search, written in the script: grid_search_cv_algorithm.R. 
This method of cross-validation results in an estimate for the degrees of freedom, and we adopt the degrees of freedom for each spline that minimized the cross-validated RMS error. 

ML = 2
fc = 6
Mc2 = 2
Mws = 10
```{r}
#set our degrees of freedom
ml.dof <- 2
fc.dof <- 6
mc2.dof <- 2
mws.dof <- 10

#get grids for predicting and plotting
ml.grid <- seq(from = range(test$ML)[1], to = range(test$ML)[2])
fc.grid <- seq(from = range(test$fc)[1], to = range(test$fc)[2])
mc2.grid <- seq(from = range(test$Mc2)[1], to = range(test$Mc2)[2])
mws.grid <- seq(from = range(test$Mws)[1], to = range(test$Mws)[2])

#use natural splines to fit the general additive model
spline.mod <- gam(Dep ~ ns(ML, df = ml.dof)
                  + ns(fc, df = fc.dof)
                  + ns(Mc2, df = mc2.dof)
                  + ns(Mws, df = mws.dof),
                  data = train)
```

Model Performance
```{r}
pred <- predict(spline.mod, newdata = test)
rmse <- rmse(test$Dep, pred)
print("RMSE for GAM using natural splines:")
print(rmse)
```
The RMSE for the GAM using natural splines with the cross-validated degrees of freedom is 3.677. This is lower than the previous OLS and Lasso models. In this instance, as the model flexibility increased from ML - Mc2 to the four-feature OLS model, to the four-feature GAM model, our ability to estimate depth improved. This provides evidence that a more flexible model is necessary to capture non-linearity in the relationship between our predictors and depth. 

Visualization
```{r}
test$resids <- test$Dep - pred
plot(test$Dep, test$resids, main = "Depth Residuals for Natural Spline GAM")
```
Again, we see the same linear relationship between the depths and residuals, indicating that although the physics-based features of the event signals are slightly predictive of depth, most of the variance is explained by the bias of the model. 

Save residuals in CSV to access later.
```{r}
write.csv(test$resids, file = "C:/Users/riley/ds_portfolio/depth_analysis/spline_model_test_resids.csv")
```
